Accelerated oblique random survival forests

### 2022-06-15 docs/index.Rmd
paper edits
grouping all the RFs into axis based or oblique should make the paper's
main hypotheses easier to test.

- $t_1 < \, \ldots \, < t_m$ are the $m$ unique event times in the training data.

@@

How do we estimate overall performance based on all 23 risk prediction tasks?

@@

    + split the corresponding data into a 50/50 train/test split
    + fit each learner to the training set
    + evaluate each learner's predictions in the testing set

@@

Step 2: Fit a hierarchical Bayesian model to analyze posterior expected performance for each learner:

>>>>

    + $T_i$ is the event time if an event occurred and last point of contact otherwise.

    + $t_1 < \, \ldots \, < t_m$ are the $m$ unique event times in the training data.

@@

To estimate overall performance differences:

@@

    + split the corresponding data into a 50/50 train/test split
    + fit each learner to the training set
    + evaluate each learner's predictions in the testing set
    + repeat 25 times

@@

Step 2: Fit a hierarchical Bayesian model to analyze posterior expected differences in performance:

### 2022-06-15 paper/jmlr/main.Rnw
paper edits
grouping all the RFs into axis based or oblique should make the paper's
main hypotheses easier to test.

@@

Consider the usual framework for survival analysis with training data $$\dataset_{\text{train}} = \left\{ (T_i, \delta_i, \bm{x}_{i}) \right\}_{i=1}^{N_{\text{train}}}.$$ Here, $T_i$ is the event time if $\delta_i=1$ and last point of contact if $\delta_i=0$, and $\bm{x}_i$ is a vector of predictors values. Assuming there are no ties, let $t_1 < \, \ldots \, < t_m$ denote the $m$ unique event times in $\dataset_{\text{train}}$. We propose to identify linear combinations of predictor variables in non-leaf nodes by applying Newton Raphson scoring to the partial likelihood function of the Cox regression model:
\begin{equation}\label{eqn:cox-partial-likelihood}
L(\bm\beta) = \prod_{i=1}^m \frac{e^{\bm{x}_{j(i)}^T \bm\beta}}{\sum_{j \in R_i} e^{\bm{x}_j^T \bm\beta}},
\end{equation}
where $R_i$ is the set of indices, $j$, with $T_j \geq t_i$ (i.e., those still at risk at time $t_i$), and $j(i)$ is the index of the observation for which an event occurred at time $t_i$. The \texttt{survival} package includes documentation that outlines how to complete this estimation procedure efficiently \citep[see][exact.nw]{therneau_survival_2022}. Briefly, a vector of estimated regression coefficients, $\hat{\beta}$, is updated in each step of the procedure based on its first derivative, $U(\hat{\beta})$, and second derivative, $H(\hat{\beta})$:
$$ \hat{\beta}^{k+1} =  \hat{\beta}^{k} + U(\hat{\beta} = \hat{\beta}^{k})\, H^{-1}(\hat{\beta} = \hat{\beta}^{k}) $$
It is vital to cycle through iterations until a convergence threshold is met for statistical inference, but it is only necessary to complete one iteration to identify coefficients for a linear combination of predictors. \citet{jaeger2019oblique} identified linear combinations using penalized regression models, which supply more flexible solutions for $\hat{\beta}$ at the cost of greater computational demand.

@@

In the current study, we consider four classes of learners: RFs, boosting ensembles, regression models, and neural networks (Table \ref{table:learners}). For RF learners, the number of observations required in terminal nodes was fixed at 10, the number of randomly selected predictors was the square root of the total number of predictors rounded to the nearest integer, and the number of trees in the ensemble was 500. For boosting and regression learners, nested cross-validation was applied to identify the number of boosting steps and the magnitude of penalization, respectively.

@@

\begin{table}[h!]
\centering
\begin{tabular}{p{2cm} | p{3cm} p{4cm} p{12cm}}
 \hline
 Learner Class & Software & Learners & Description \\ [0.5ex]
 \hline\hline
 \multicolumn{3}{l}{\textit{Random Survival Forests}}\\
 \hline\hline
 Standard & \href{https://www.randomforestsrc.org/index.html}{\texttt{RandomForestSRC}} & \texttt{rfsrc-standard} & Axis based survival trees following Leo Breiman's original random forest algorithm, with cut-points selected to maximize a log-rank statistic.  \\ \hline
 Oblique & \href{https://CRAN.R-project.org/package=obliqueRSF}{\texttt{obliqueRSF}} \newline \href{https://bcjaeger.github.io/aorsf/}{\texttt{aorsf}} &
 \texttt{obliqueRSF-net} \newline
 \texttt{aorsf-net} \newline
 \texttt{aorsf-cph($i=1$)} \newline
 \texttt{aorsf-cph($i \leq 15$)} \newline
 \texttt{aorsf-extratrees} &
 Oblique survival trees following Leo Breiman's random forest algorithm. Linear combinations of inputs are derived using \texttt{glmnet} in \texttt{obliqueRSF-net} and \texttt{aorsf-net}, using Newton Raphson scoring for the Cox partial likelihood function in \texttt{aorsf-cph($i=1$)} and \texttt{aorsf-cph($i \leq 15$)}, and chosen randomly from a uniform distribution in \texttt{aorsf-extratrees}. Cut-points are selected to maximize a log-rank statistic. \\ \hline
 Extremely \,Randomized & \href{https://CRAN.R-project.org/package=ranger}{\texttt{ranger}} & \texttt{ranger-extratrees} & Axis-based survival trees grown with randomly selected features and cut-points\\ \hline
 Conditional \,Inference & \href{http://party.r-forge.r-project.org/}{\texttt{party}} & \texttt{party-cif} & Axis based survival trees grown using unbiased recursive partitioning.  \\
 \hline\hline
 \multicolumn{3}{l}{\textit{Boosting ensembles}}\\
 \hline\hline
 Trees & \href{https://xgboost.readthedocs.io/en/stable/#}{\texttt{xgboost}} & \texttt{xgboost-cox} &  The Cox partial likelihood function is maximized additively with decision trees. Nested cross validation (5 folds) is applied to tune the number of trees grown.  \\
 Models & \href{https://xgboost.readthedocs.io/en/stable/#}{\texttt{xgboost}} & \texttt{xgboost-aft} & The accelerated failure time likelihood function is maximized additively with decision trees. Nested cross validation (5 folds) is applied to tune the number of trees grown.  \\
 \hline\hline
 \multicolumn{3}{l}{\textit{Regression models}}\\
 \hline\hline
 Cox Net & \texttt{glmnet} & \texttt{glmnet-cox} & The Cox model is fit using an elastic net penalty. Nested cross validation (5 folds) is applied to tune penalty terms.\\
 \hline\hline
 \multicolumn{3}{l}{\textit{Neural networks}}\\
 \hline\hline
 Cox Time & \href{https://raphaels1.github.io/survivalmodels/}{\texttt{survivalmodels}} & \texttt{nn-cox} & A neural network based on the proportional hazards model with time-varying effects  \\
 \hline
\end{tabular}
\caption{Learning algorithms assessed in numeric studies}
\label{table:learners}
\end{table}

>>>>

Consider the usual framework for survival analysis with training data $$\dataset_{\text{train}} = \left\{ (T_i, \delta_i, \bm{x}_{i}) \right\}_{i=1}^{N_{\text{train}}}.$$ Here, $T_i$ is the event time if $\delta_i=1$ and last point of contact if $\delta_i=0$, and $\bm{x}_i$ is a vector of predictors values. Assuming there are no ties, let $t_1 < \, \ldots \, < t_m$ denote the $m$ unique event times in $\dataset_{\text{train}}$.

@@

We propose to identify linear combinations of predictor variables in non-leaf nodes by applying Newton Raphson scoring to the partial likelihood function of the Cox regression model:
\begin{equation}\label{eqn:cox-partial-likelihood}
L(\bm\beta) = \prod_{i=1}^m \frac{e^{\bm{x}_{j(i)}^T \bm\beta}}{\sum_{j \in R_i} e^{\bm{x}_j^T \bm\beta}},
\end{equation}
where $R_i$ is the set of indices, $j$, with $T_j \geq t_i$ (i.e., those still at risk at time $t_i$), and $j(i)$ is the index of the observation for which an event occurred at time $t_i$. The \texttt{survival} package includes documentation that outlines how to complete this estimation procedure efficiently \citep[see][exact.nw]{therneau_survival_2022}. Briefly, a vector of estimated regression coefficients, $\hat{\beta}$, is updated in each step of the procedure based on its first derivative, $U(\hat{\beta})$, and second derivative, $H(\hat{\beta})$:
$$ \hat{\beta}^{k+1} =  \hat{\beta}^{k} + U(\hat{\beta} = \hat{\beta}^{k})\, H^{-1}(\hat{\beta} = \hat{\beta}^{k}) $$
It is vital to cycle through iterations until a convergence threshold is met for statistical inference, but it is only necessary to complete one iteration to identify coefficients for a linear combination of predictors. \citet{jaeger2019oblique} identified linear combinations using penalized regression models, which supply more flexible solutions for $\hat{\beta}$ at the cost of greater computational demand.

@@

We consider four classes of learners: RFs, boosting ensembles, regression models, and neural networks (Table \ref{table:learners}). For RF learners, the number of observations required in terminal nodes was fixed at 10, the number of randomly selected predictors was the square root of the total number of predictors rounded to the nearest integer, and the number of trees in the ensemble was 500. For boosting, regression, and neural network learners, nested cross-validation was applied to tune each model, \ie, assess multiple possible values for relevant model parameters and use the values that optimize the model's estimated prediction accuracy. Specifically, tuning for boosting models included identifying the number of steps to complete. For regression models, tuning was used to identify the magnitude of penalization. For neural networks, the number and density of layers was tuned.

@@

\begin{table}[h!]
\centering
\begin{tabular}{p{2cm} | p{3cm} p{4cm} p{12cm}}
 \hline
 Learner Class & Software & Learners & Description \\ [0.5ex]
 \hline\hline
 \multicolumn{3}{l}{\textit{Random Survival Forests}}\\
 \hline\hline

 Axis based &

 \href{https://www.randomforestsrc.org/index.html}{\texttt{RandomForestSRC}} \newline
 \href{https://CRAN.R-project.org/package=ranger}{\texttt{ranger}} \newline
 \href{http://party.r-forge.r-project.org/}{\texttt{party}} \newline
 \href{https://github.com/whcsu/rotsf}{\texttt{rotsf}} \newline
 \href{https://github.com/whcsu/rsfse}{\texttt{rsfse}} &

 \texttt{rsf-standard} \newline
 \texttt{rsf-extratrees} \newline
 \texttt{cif-standard} \newline
 \texttt{cif-rotate} \newline
 \texttt{cif-spacextend} &

 \texttt{rsf-standard} grows survival trees following Leo Breiman's original random forest algorithm with variables and cut-points selected to maximize a log-rank statistic. \texttt{rsf-extratrees} grows survival trees with randomly selected features and cut-points. \texttt{cif-standard} uses the framework of conditional inference to grow survival trees. \texttt{cif-rotate} extends \texttt{cif-standard} by applying principal component analysis to random subsets of data prior to growing each survival tree. \texttt{cif-spacextend} derives new predictors for each tree in the ensemble, separately. \\ \hline

 Oblique &

 \href{https://CRAN.R-project.org/package=obliqueRSF}{\texttt{obliqueRSF}} \newline
 \href{https://bcjaeger.github.io/aorsf/}{\texttt{aorsf}} &

 \texttt{obliqueRSF-net} \newline
 \texttt{aorsf-net} \newline
 \texttt{aorsf-fast} \newline
 \texttt{aorsf-cph} \newline
 \texttt{aorsf-extratrees} &

 Oblique survival trees following Leo Breiman's random forest algorithm. Linear combinations of inputs are derived using \texttt{glmnet} in \texttt{obliqueRSF-net} and \texttt{aorsf-net}, using Newton Raphson scoring for the Cox partial likelihood function in \texttt{aorsf-fast} and \texttt{aorsf-cph}, and chosen randomly from a uniform distribution in \texttt{aorsf-extratrees}. Cut-points are selected to maximize a log-rank statistic. \\

 \hline\hline
 \multicolumn{3}{l}{\textit{Boosting ensembles}}\\
 \hline\hline

 Trees &

 \href{https://xgboost.readthedocs.io/en/stable/#}{\texttt{xgboost}} &

 \texttt{xgboost-cox} \newline
 \texttt{xgboost-aft} &

 \texttt{xgboost-cox} maximizes the Cox partial likelihood function, whereas \texttt{xgboost-aft} maximizes the accelerated failure time likelihood function. Nested cross validation (5 folds) is applied to tune the number of trees grown, the minimum number of observations in a leaf node was 10, the maximum depth of trees was 6, and $\sqrt{p}$ variables were considered randomly for each tree split, where $p$ is the total number of predictors. \\
 \hline\hline
 \multicolumn{3}{l}{\textit{Regression models}}\\
 \hline\hline
 Cox Net & \texttt{glmnet} & \texttt{glmnet-cox} & The Cox proportional hazards model is fit using an elastic net penalty. Nested cross validation (5 folds) is applied to tune penalty terms.\\
 \hline\hline
 \multicolumn{3}{l}{\textit{Neural networks}}\\
 \hline\hline
 Cox Time & \href{https://raphaels1.github.io/survivalmodels/}{\texttt{survivalmodels}} & \texttt{nn-cox} & A neural network based on the proportional hazards model with time-varying effects. Nested cross-validation was applied to select the number of layers (from 1 to 8), the number of nodes in each layer (from $\sqrt{p}$/2 to $\sqrt{p}$), and the number of epochs to complete (up to 500). A drop-out rate of 10\% was applied during training.   \\
 \hline
\end{tabular}
\caption{Learning algorithms assessed in numeric studies}
\label{table:learners}
\end{table}
